Extrinsic Type Inference (DRAFT 2025-06-09)

### 2025-07-09 extrinsic_type_inference_paper/paper.tex
update
add positive substitution requirement

\documentclass[acmsmall]{acmart}
% \documentclass[letterpaper]{llncs}
% \usepackage[letterpaper, margin=1.5in]{geometry}

\usepackage{multicol}
\usepackage{mathpartir}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{nccmath}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage[scaled]{beramono}
\usepackage[T1]{fontenc}
\usepackage[dvipsnames]{xcolor}
\usepackage{array}
\usepackage{wasysym}

@@

\usepackage[table]{xcolor}

@@

\theoremstyle{definition}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{definition}{Definition}[section]
\newtheorem{property}{Property}[section]
% \newtheorem{proof}{Proof}[section]

@@

\title{Extrinsic Type Inference (DRAFT 2025-06-08)}
\author{Thomas Logan}

@@

\newcommand{\examp}[1]{{\nopad \footnotesize 
\vspace{-1em}
\[
\begin{array}[t]{@{}l}
#1
\end{array}
\]
\vspace{-1em}
}}

@@

\TODO{add requirements on factoring, packing, polar sub and tidying, and negatble checks}

@@

In order to maintain HM's implicit first-order types in the context
of higher-order types, other techniques were explored.  
These techniques include boxed polymorphism \cite{}, ML$^F$ \cite{}, and various other systems \cite{}. 
Unfortunately, it's not clear what parametric type should be inferred for \code{f}.
For instance, \code{forall 'a . 'a -> 'a}, \code{forall 'a . 'a -> 'a list}, \code{forall 'a . 'a -> 'a option}, 
or an infinite number of
other parametric types could be used, but none is weaker than all the rest.
Thus, these systems with impredicative parametric polymorphism rely on annotations of parameters to
determine second order parametric types, which constitutes intrinsic typing of
of functions. 
Since there is no parameter type that is weaker than all the rest,
merely relying on intrinsic second order parametric types 
significantly restricts the reusability of expressions.
Additionally, many of these techniques require explicit instantiation to use parametrically typed
expressions, which can be seen as an intrinsic restriction that 
delays pruning the control-flow until a particular usage.

>>>>

\documentclass[table,dvipsnames,acmsmall]{acmart}
% \documentclass[letterpaper]{llncs}
% \usepackage[letterpaper, margin=1.5in]{geometry}

\usepackage{multicol}
\usepackage{mathpartir}
\usepackage{hyperref}
\usepackage{mathtools}
\usepackage{amsmath}
\usepackage{nccmath}
\usepackage{stmaryrd}
\usepackage{listings}
\usepackage[scaled]{beramono}
\usepackage[T1]{fontenc}
\usepackage[table,dvipsnames]{xcolor}
\usepackage{array}
\usepackage{wasysym}

@@

@@

\theoremstyle{definition}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{requirement}{Requirement}[section]
\newtheorem{definition}{Definition}[section]
\newtheorem{property}{Property}[section]
% \newtheorem{proof}{Proof}[section]

@@

\title{Extrinsic Type Inference (DRAFT 2025-06-09)}
\author{Thomas Logan}

@@

\newcommand{\examp}[1]{{\nopad \footnotesize 
\[
\begin{array}[t]{@{}l}
#1
\end{array}
\]
}}

@@

\begin{requirement}
  \label{req:positive_substitution_soundness}
  \emph{positive substitution soundness}
  \small
  \nopad
  \begin{mathpar}
    \inferrule {
      \alpha \notin \textbf{ftv}(\tau[\alpha \slash \tau_r]^+)
    } {
      \delta \satisfies \tau_l \subtypes \tau_r
      \implies
      \delta\ \alpha \slash \tau_l \satisfies \tau \subtypes \tau[\alpha \slash \tau_r]^+
    }
  \end{mathpar}
\end{requirement}

\TODO{say what it means simply}
This property is necessary so that our proof typing rules can use positive substitution
as a proxy reasoning under the induction hypothesis.

\TODO{add requirements on factoring, packing, and tidying, and negatble checks}

@@

In order to maintain HM's implicit first-order types in the context
of higher-order types, other techniques were explored.  
These techniques include boxed polymorphism \cite{}, ML$^F$ \cite{}, and various other systems \cite{}. 
Unfortunately, it's not clear what parametric type should be inferred for \code{f}.
There are an infinite number of
parametric types that could be used, but none is weaker than all the rest.

\examp{
  \J{forall 'a . 'a -> 'a}
  \\
  \J{forall 'a . 'a -> 'a list}
  \\
  \J{forall 'a . 'a -> 'a option}
}

Thus, these systems with impredicative parametric polymorphism rely on annotations of parameters to
determine second order parametric types, which constitutes intrinsic typing of
of functions. 
Since there is no parameter type that is weaker than all the rest,
merely relying on intrinsic second order parametric types 
significantly restricts the reusability of expressions.
Additionally, many of these techniques require explicit instantiation to use parametrically typed
expressions, which can be seen as an intrinsic restriction that 
delays pruning the control-flow until a particular usage.
